---
title: "520- Project"
author: "Ritika Lama"
format: pdf
editor: visual
---

## Dataset

```{r}
library(tidyverse)
gaming <- read_csv("gaming_data.csv")
names(gaming)
```

## Clean the dataset and summarise

```{r}
gaming_clean <- gaming %>%
  mutate(
    LastPurchaseDate = suppressWarnings(ymd(LastPurchaseDate)),
    InAppPurchaseAmount = if_else(is.na(InAppPurchaseAmount), 0, InAppPurchaseAmount),
    PaymentMethod = if_else(is.na(PaymentMethod) & InAppPurchaseAmount == 0, "No Purchase", PaymentMethod),
    Age = if_else(is.na(Age), median(Age, na.rm = TRUE), Age),
    Gender = replace_na(Gender, "Unknown"),
    Country = replace_na(Country, "Unknown"),
    Device = replace_na(Device, "Unknown"),
    GameGenre = replace_na(GameGenre, "Unknown")
  )

max_purchase_date <- max(gaming_clean$LastPurchaseDate, na.rm = TRUE)

gaming_clean <- gaming_clean %>%
  mutate(
    RecencyDays = as.numeric(max_purchase_date - LastPurchaseDate),
    TotalPlaytime = SessionCount * AverageSessionLength,
    LogSpend = log1p(InAppPurchaseAmount),
    IsPayer = if_else(InAppPurchaseAmount > 0, 1, 0)
  )

write.csv(gaming_clean, "gaming_clean.csv")

head(gaming_clean)
str(gaming_clean)
names(gaming_clean)
colSums(is.na(gaming_clean))
summary(gaming_clean)
```

```{r}
# -----------------------------
# Graph 1: Distribution of Spending
# -----------------------------
ggplot(gaming_clean, aes(x = InAppPurchaseAmount)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of In-App Purchase Spending",
    x = "In-App Purchase Amount ($)",
    y = "Number of Players"
  ) +
  theme_minimal()

# -----------------------------
# Graph 2: Session Count vs Spending
# -----------------------------
ggplot(gaming_clean, aes(x = SessionCount, y = InAppPurchaseAmount)) +
  geom_point(alpha = 0.4, color = "darkgreen") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(
    title = "Session Count vs In-App Spending",
    x = "Session Count",
    y = "In-App Purchase Amount ($)"
  ) +
  theme_minimal()

# -----------------------------
# Graph 3: Average Session Length vs Spending
# -----------------------------
ggplot(gaming_clean, aes(x = AverageSessionLength, y = InAppPurchaseAmount)) +
  geom_point(alpha = 0.4, color = "purple") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(
    title = "Average Session Length vs Spending",
    x = "Average Session Length (minutes)",
    y = "In-App Purchase Amount ($)"
  ) +
  theme_minimal()

# -----------------------------
# Graph 4: Average Spending by Game Genre
# -----------------------------
gaming_clean %>%
  group_by(GameGenre) %>%
  summarise(AverageSpending = mean(InAppPurchaseAmount, na.rm = TRUE)) %>%
  arrange(desc(AverageSpending)) %>%
  ggplot(aes(x = reorder(GameGenre, AverageSpending), y = AverageSpending)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  labs(
    title = "Average Spending by Game Genre",
    x = "Game Genre",
    y = "Average Spending ($)"
  ) +
  theme_minimal()

# -----------------------------
# Graph 5: Boxplot of Spending by Genre
# -----------------------------
ggplot(gaming_clean, aes(x = GameGenre, y = InAppPurchaseAmount)) +
  geom_boxplot(fill = "lightgreen") +
  coord_flip() +
  labs(
    title = "Spending Distribution by Game Genre",
    x = "Game Genre",
    y = "In-App Purchase Amount ($)"
  ) +
  theme_minimal()



```

```{r}
# ---------------------------
# Modeling dataset
# ---------------------------
model_df <- gaming_clean %>%
  select(
    LogSpend, InAppPurchaseAmount, Age, Gender, Country, Device, GameGenre,
    SessionCount, AverageSessionLength, FirstPurchaseDaysAfterInstall,
    PaymentMethod, RecencyDays, TotalPlaytime
  ) %>%
  drop_na(LogSpend, Age, Gender, Country, Device, GameGenre,
          SessionCount, AverageSessionLength, PaymentMethod, RecencyDays, TotalPlaytime)

#Split training and testing
set.seed(42)
train_index <- sample(seq_len(nrow(model_df)), size = 0.8 * nrow(model_df))
train <- model_df[train_index, ]
test <- model_df[-train_index, ]
```

## K-means Player Segmentation (Minnow, Dolphin, Whale)

```{r}
set.seed(42)
# Use payer + engagement features for clustering
cluster_df <- gaming_clean %>%
  select(UserID, InAppPurchaseAmount, SessionCount, AverageSessionLength, TotalPlaytime) %>%
  drop_na() %>%
  filter(InAppPurchaseAmount > 0)

cluster_scaled <- cluster_df %>%
  select(InAppPurchaseAmount, SessionCount, AverageSessionLength, TotalPlaytime) %>%
  scale()

kmeans_model <- kmeans(cluster_scaled, centers = 3, nstart = 30)

cluster_results <- cluster_df %>%
  mutate(cluster = kmeans_model$cluster) %>%
  group_by(cluster) %>%
  mutate(cluster_avg_spend = mean(InAppPurchaseAmount, na.rm = TRUE)) %>%
  ungroup()

# Rank clusters by average spending and label as Minnow/Dolphin/Whale
cluster_rank <- cluster_results %>%
  distinct(cluster, cluster_avg_spend) %>%
  arrange(cluster_avg_spend) %>%
  mutate(PlayerType = c("Minnow", "Dolphin", "Whale"))

cluster_results <- cluster_results %>%
  left_join(cluster_rank, by = c("cluster", "cluster_avg_spend"))

cluster_rank

cluster_results %>%
  group_by(PlayerType) %>%
  summarise(
    Players = n(),
    AvgSpend = mean(InAppPurchaseAmount, na.rm = TRUE),
    AvgSessionCount = mean(SessionCount, na.rm = TRUE),
    AvgSessionLength = mean(AverageSessionLength, na.rm = TRUE),
    AvgTotalPlaytime = mean(TotalPlaytime, na.rm = TRUE)
  ) %>%
  arrange(match(PlayerType, c("Minnow", "Dolphin", "Whale")))
```


```{r}
# 2D cluster map using PCA (similar to classic K-means cluster visualization)
pca_fit <- prcomp(cluster_scaled)

cluster_plot_df <- as_tibble(pca_fit$x[, 1:2]) %>%
  setNames(c("PC1", "PC2")) %>%
  mutate(PlayerType = cluster_results$PlayerType)

cluster_centers <- cluster_plot_df %>%
  group_by(PlayerType) %>%
  summarise(PC1 = mean(PC1), PC2 = mean(PC2), .groups = "drop")

ggplot(cluster_plot_df, aes(x = PC1, y = PC2, color = PlayerType)) +
  geom_point(alpha = 0.75, size = 2) +
  geom_point(
    data = cluster_centers,
    aes(x = PC1, y = PC2),
    color = "black",
    size = 5,
    shape = 4,
    stroke = 1.2,
    inherit.aes = FALSE
  ) +
  labs(
    title = "Estimated Number of Clusters: 3",
    x = "Principal Component 1",
    y = "Principal Component 2",
    color = "Segment"
  ) +
  theme_minimal()
```

```{r}
# Cluster sizes (how many players in each segment)
cluster_results %>%
  count(PlayerType) %>%
  mutate(PlayerType = factor(PlayerType, levels = c("Minnow", "Dolphin", "Whale"))) %>%
  ggplot(aes(x = PlayerType, y = n, fill = PlayerType)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  labs(
    title = "Player Count by K-means Segment",
    x = "Segment",
    y = "Number of Players"
  ) +
  theme_minimal()
```


```{r}
# Segment spending profile
cluster_results %>%
  group_by(PlayerType) %>%
  summarise(AvgSpend = mean(InAppPurchaseAmount, na.rm = TRUE)) %>%
  mutate(PlayerType = factor(PlayerType, levels = c("Minnow", "Dolphin", "Whale"))) %>%
  ggplot(aes(x = PlayerType, y = AvgSpend, fill = PlayerType)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  labs(
    title = "Average Spending by Segment",
    x = "Segment",
    y = "Average In-App Purchase Amount ($)"
  ) +
  theme_minimal()
```

```{r}
# Distribution of spending within each cluster
cluster_results %>%
  mutate(PlayerType = factor(PlayerType, levels = c("Minnow", "Dolphin", "Whale"))) %>%
  ggplot(aes(x = PlayerType, y = InAppPurchaseAmount, fill = PlayerType)) +
  geom_boxplot(show.legend = FALSE, alpha = 0.8) +
  labs(
    title = "Spending Distribution Within Each Segment",
    x = "Segment",
    y = "In-App Purchase Amount ($)"
  ) +
  theme_minimal()
```


## Logistic Model: High vs Low Spender

After segmentation, we model whether a player is a high spender or not using behavior-related predictors.

```{r}
# Binary target: high spender if spending >= 75th percentile
threshold <- quantile(gaming_clean$InAppPurchaseAmount, 0.75, na.rm = TRUE)

class_df <- gaming_clean %>%
  mutate(
    HighSpender = if_else(InAppPurchaseAmount >= threshold, 1, 0)
  ) %>%
  select(
    HighSpender, Age, SessionCount, AverageSessionLength,
    FirstPurchaseDaysAfterInstall, RecencyDays, TotalPlaytime
  ) %>%
  drop_na()

set.seed(42)
train_index_cls <- sample(seq_len(nrow(class_df)), size = 0.8 * nrow(class_df))
train_cls <- class_df[train_index_cls, ]
test_cls <- class_df[-train_index_cls, ]

# Full logistic model with behavior variables
glm_full <- glm(
  HighSpender ~ Age + SessionCount + AverageSessionLength +
    FirstPurchaseDaysAfterInstall + RecencyDays + TotalPlaytime,
  data = train_cls,
  family = binomial()
)

glm_summary <- summary(glm_full)
glm_summary

# Keep only significant variables (p < 0.05); fallback if none
coefs <- coef(glm_summary)
sig_terms <- rownames(coefs)[coefs[, "Pr(>|z|)"] < 0.05]
sig_terms <- setdiff(sig_terms, "(Intercept)")
if (length(sig_terms) == 0) {
  sig_terms <- c("SessionCount", "TotalPlaytime")
}

glm_formula <- as.formula(
  paste("HighSpender ~", paste(sig_terms, collapse = " + "))
)

glm_sig <- glm(glm_formula, data = train_cls, family = binomial())
summary(glm_sig)
```


```{r}
# Classification performance on test set
glm_prob <- predict(glm_sig, newdata = test_cls, type = "response")
glm_pred <- if_else(glm_prob >= 0.5, 1, 0)

glm_cm <- table(Predicted = glm_pred, Actual = test_cls$HighSpender)
glm_cm

glm_accuracy <- mean(glm_pred == test_cls$HighSpender)
glm_accuracy
```


## Random Forest Classification (High vs Low Spender)

To compare with a non-linear model, we fit a random forest and inspect which variables contribute most to classification.

```{r}
library(randomForest)
set.seed(42)
rf_class_model <- randomForest(
  as.factor(HighSpender) ~ Age + SessionCount + AverageSessionLength +
    FirstPurchaseDaysAfterInstall + RecencyDays + TotalPlaytime,
  data = train_cls,
  ntree = 500,
  importance = TRUE
)

rf_class_model
importance(rf_class_model)
varImpPlot(rf_class_model)
```

The following test results show classification accuracy and confusion matrix for random forest.

```{r}
# Random forest test performance
rf_pred <- predict(rf_class_model, newdata = test_cls)
rf_cm <- table(Predicted = rf_pred, Actual = as.factor(test_cls$HighSpender))
rf_cm

rf_accuracy <- mean(rf_pred == as.factor(test_cls$HighSpender))
rf_accuracy
```

## Model Comparison (Logistic vs Random Forest)

This final table places both models side by side using the same metrics (Accuracy, Precision, Recall, F1).

```{r}
# Convert predictions to factors with consistent levels
actual_factor <- factor(test_cls$HighSpender, levels = c(0, 1))
glm_factor <- factor(glm_pred, levels = c(0, 1))
rf_factor <- factor(as.numeric(as.character(rf_pred)), levels = c(0, 1))

# Helper function for binary classification metrics
calc_metrics <- function(pred, actual) {
  cm <- table(Predicted = pred, Actual = actual)
  
  tp <- ifelse(!is.na(cm["1", "1"]), cm["1", "1"], 0)
  tn <- ifelse(!is.na(cm["0", "0"]), cm["0", "0"], 0)
  fp <- ifelse(!is.na(cm["1", "0"]), cm["1", "0"], 0)
  fn <- ifelse(!is.na(cm["0", "1"]), cm["0", "1"], 0)
  
  accuracy <- (tp + tn) / sum(cm)
  precision <- ifelse((tp + fp) == 0, 0, tp / (tp + fp))
  recall <- ifelse((tp + fn) == 0, 0, tp / (tp + fn))
  f1 <- ifelse((precision + recall) == 0, 0, 2 * precision * recall / (precision + recall))
  
  tibble(
    Accuracy = round(accuracy, 3),
    Precision = round(precision, 3),
    Recall = round(recall, 3),
    F1 = round(f1, 3)
  )
}

comparison_table <- bind_rows(
  calc_metrics(glm_factor, actual_factor) %>% mutate(Model = "Logistic Regression"),
  calc_metrics(rf_factor, actual_factor) %>% mutate(Model = "Random Forest")
) %>%
  select(Model, Accuracy, Precision, Recall, F1)

comparison_table
```
